Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
fix gaps

Building on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany at the federal‐state level, this paper (1) identifies and characterizes the key challenges involved in assessing segmentation accuracy over expansive, heterogeneous terrains––including variability in canopy structure, imaging conditions, and regional ecological gradients––and (2) provides concrete methodological recommendations to address these challenges. By coupling a rigorous evaluation of terrain‐specific segmentation limitations with these targeted recommendations, we aim to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales required by global environmental challenges.

@@

![a. Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
b. Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees. A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.]("./figures/figure1.png")

@@

![Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.]("./figures/figure2.png")

>>>>

Building on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany at the federal‐state level, this paper (1) identifies and characterizes the key challenges involved in assessing segmentation accuracy over expansive, heterogeneous terrains–including variability in canopy structure, imaging conditions, and regional ecological gradients–and (2) provides concrete methodological recommendations to address these challenges. By coupling a rigorous evaluation of terrain‐specific segmentation limitations with these targeted recommendations, we aim to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales required by global environmental challenges.

@@

::: {.column width="100%"}
```{=latex}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/figure1.png}
\caption{a. Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
b. Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees. A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.}
\end{figure}

@@

::: {.column width="100%"}
```{=latex}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/figure2.png}
\caption{Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue I\& Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.}
\end{figure}
